Within chert: a multi-technique mineral and geochemical approach to the study of chert of southwestern Iberia

### 2025-04-24 analysis/paper/paper.qmd
Desktop session. Paper render and improvements.

---
title: "Within chert: a multi-technique mineral and geochemical approach to the study of chert of southwestern Iberia"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3

@@

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(ggpmisc)
library(magick)
library(ggpubr)

@@

# Read grouped supporting data
xrd_perc <- read_csv('analysis/data/raw_data/xrd_data/xrd_goals.csv') # File with component percentage of all samples for annotations
xrd_marks <- read_csv('analysis/data/raw_data/xrd_data/xrd_peaks.csv') # File with peak data and d spacing for plot labels

# Read individual xrd files
sp15_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP15_xrd.csv')
sp58_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP58_xrd.csv')
sp69_xrd <- read_csv('analysis/data/raw_data/xrd_data/SP69_xrd.csv')

@@

M18_7 <- read_csv('analysis/data/raw_data/sem_data/M18_7.csv')
M18_8 <- read_csv('analysis/data/raw_data/sem_data/M18_8.csv')
M4_3 <- read_csv('analysis/data/raw_data/sem_data/M4_3.csv')

@@

ggsave("M4_3_graph.png", width = 5, height = 3.75, units = "in", ,bg = "white" , dpi = 300, path = "analysis/figures/SOM/Graphs")

@@

pxrf_raw <- read_csv('analysis/data/raw_data/pxrf_data/pxrf_data_dec.csv')

@@

# Read table
tbl_samples <- read_csv('analysis/data/raw_data/geochem_samples_n.csv')
# Turn NAs into empty strings
tbl_samples <- tbl_samples %>%
  mutate(across(everything(), ~ replace(.x, is.na(.x), "")))

@@

# Read table
xrd_results_tbl <- read_csv('analysis/data/raw_data/xrd_results.csv')
# Turn NAs into empty strings
xrd_results_tbl <- xrd_results_tbl %>%
  mutate(across(everything(), ~ replace(.x, is.na(.x), "")))

@@

```{r}
#| label: fig-xrd-15
#| fig-cap: "Plot of minerals identified through XRD in the Lower Jurassic (Carixian) chert sample from the outcrop Praia do Belixe (SP15_PBX). The rhombus shape signifies the main peaks key to identify each mineral. Labels represent the d spacing values of key mineral peaks. Quantifications presented in the supporting table were obtained using Rietveld refinement."

@@

```{r}
#| label: fig-skeletal
#| fig-cap: "A) Skeletal remains from sample SP16_BLX, from the Lower Jurassic outcrops of western Algarve. B) Detail of skeletal remains composed of calcium (B1), phosphorus (B2) and a small peak of sulphur (B3). The skeletal remains seem to be oxidized, with small peaks of iron within (B4)."

knitr::include_graphics(here::here("figures/within-chert/M3_3_map1.png"))

@@

```{r}
#| label: fig-oval-fossils
#| fig-cap: "A-C) General view and details from an archaeological sample from local chert type (T4). A) general view of a concentration of oval fossils. B) general view of a concentration of oval fossils and other inclusions. C) detail of an irregular inclusion composed of titanium (C1) and an oval skeletal remains composed mainly of calcium (C2), phosphorus (C3) and a small peak of sulphur."

knitr::include_graphics(here::here("figures/within-chert/M15_4.png"))

@@

```{r}
#| label: fig-t6-general
#| fig-cap: "A-D) General view and details from an archaeological sample from non-local chert type (T6). Dots represent the measured spectrum and shown in the corresponding graph. A) general view of the sample with uncommon inclusions. B) detail of an inclusion seen in A, composed mostly of iron and small peaks of calcium and phosphorus, making this inclusion a possible oxidised skeletal remains of a fossil (1). C) general view of the sample with no visible inclusions. D) detail of a small and irregular iron inclusion (2)."

knitr::include_graphics(here::here("figures/within-chert/M18_7_M18_8_graphs.png"))

@@

```{r}
#| label: fig-framboids
#| fig-cap: "A-B) Detail from the geological sample SP59_JOR from the Upper Jurassic outcrops (Jordana formation) of eastern Algarve. A) detailed view of a spherical pyrite framboid composed of iron (A1) and sulphur (A2). B) detailed view of an inclusion composed of iron (B1) and sulphur (B2)."

knitr::include_graphics(here::here("figures/within-chert/M10_5_map1_map2.png"))

@@

```{r}
#| label: fig-calcite
#| fig-cap: "General view of calcite inclusions (light grey) from the geological sample SP69_MAR from the Upper Jurassic outcrops of western Algarve, composed mainly of calcium and silica (1). Dot represents the measured spectrum and is shown in the corresponding graph."

knitr::include_graphics(here::here("figures/within-chert/M4_3_graphs.png"))

@@

When comparing the regional Algarve cherts (fig. ++ a), we see that the two first dimensions explain a combined `r 34.6+26.3`% variability in the results, with dimension 1 explaining 34.6% and dimension 2 explaining 26.3%. By observing the scatterplots, it is noticeable that the Lower Jurassic cherts (LJW) are highly variable when compared to the other cherts, with samples which display higher positive values in dimension 1 (Dim1) and dimension 2 (Dim2). However, when focusing on the concentration of LJW results, they seem to fit within the range shown by Middle Jurassic and Upper Jurassic eastern cherts. In comparison, Upper Jurassic cherts from the western basin show higher positive values in dimension 2, which is explained by the higher values of CaO.

The differences between the local Lower Jurassic cherts and non-local cherts are more marked. Fig ++ b represents the correlation between elements and Lower Jurassic cherts, central Portugal Cretaceous cherts and southern Spain Upper Jurassic cherts. Dimension 1 explains \~57% of the variability, while dimension 2 explains \~29%, with a combined percentage of `r 56.6+28.6`%. Once again, it is clear that Jurassic cherts, both from southwestern Portugal and southern Spain, show variability, although they are mostly located in the center/bottom of dimension 2 and center of dimension 1. When compared to these values, central Portugal Cretaceous cherts are located in the negative axis of dimension 1 and show little variability.

A similar pattern emerges when applying a PCA using the same elements but comparing them to archaeological, non-local chert T6, previously attributed to central Portugal (Fig. ++ c). In this PCA, dimension 1 explains \~52% of the variability and dimension 2 \~30%, with a combined percentage of `r 52.2+30.1`%. Similarly to geological Cretaceous samples from central Portugal, T6 samples are also concentrated in the negative side of dimension 1, with a clear overlap between the two.

As observed in Fig. ++ b, Fig. ++ d shows a concentration of all groups in the centre/bottom of dimension 2 (28.7%) and centre of dimension 1 (58.7%), with the first two dimensions explaining a combined percentage of `r 58.7+28.7`% of the variability. The archaeological local cherts T2 to T5 and non-local type T7 (previously attributed to the Upper Jurassic cherts of southern Spain) are congruent with the corresponding geological sources, although with clear decreased variability.

>>>>

---
title: "Within chert: a multi-technique mineral and geochemical approach to the study of chert of southwestern Iberia"
author: Joana Belmiro
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3

@@

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(ggpmisc)
library(magick)
library(ggpubr)
library(here)

@@

# Read grouped supporting data
xrd_perc <- read_csv(here::here('analysis/data/raw_data/xrd_data/xrd_goals.csv')) # File with component percentage of all samples for annotations
xrd_marks <- read_csv(here::here('analysis/data/raw_data/xrd_data/xrd_peaks.csv')) # File with peak data and d spacing for plot labels

# Read individual xrd files
sp15_xrd <- read_csv(here::here('analysis/data/raw_data/xrd_data/SP15_xrd.csv'))
sp58_xrd <- read_csv(here::here('analysis/data/raw_data/xrd_data/SP58_xrd.csv'))
sp69_xrd <- read_csv(here::here('analysis/data/raw_data/xrd_data/SP69_xrd.csv'))

@@

M18_7 <- read_csv(here::here('analysis/data/raw_data/sem_data/M18_7.csv'))
M18_8 <- read_csv(here::here('analysis/data/raw_data/sem_data/M18_8.csv'))
M4_3 <- read_csv(here::here('analysis/data/raw_data/sem_data/M4_3.csv'))

@@

# ggsave("M4_3_graph.png", width = 5, height = 3.75, units = "in", ,bg = "white" , dpi = 300, path = "analysis/figures/SOM/Graphs")

@@

pxrf_raw <- read_csv(here::here('analysis/data/raw_data/pxrf_data/pxrf_data_dec.csv'))

@@

# Read table
tbl_samples <- read_csv(here::here('analysis/data/raw_data/geochem_samples_n.csv'))
# Turn NAs into empty strings
tbl_samples <- tbl_samples %>%
  mutate(across(everything(), ~ replace(.x, is.na(.x), "")))

@@

# Read table
xrd_results_tbl <- read_csv(here::here('analysis/data/raw_data/xrd_results.csv'))
# Turn NAs into empty strings
xrd_results_tbl <- xrd_results_tbl %>%
  mutate(across(everything(), ~ replace(.x, is.na(.x), "")))

@@

```{r}
#| label: fig-xrd-15
#| fig-cap: "Plot of minerals identified through XRD in the Lower Jurassic (Carixian) chert sample from the outcrop Praia do Belixe (SP15_PBX). The rhombus shape signifies the main peaks key to identify each mineral. Labels represent the d spacing values of key mineral peaks. Quantifications presented in the supporting table were obtained using Rietveld refinement."
#| fig-width: 6
#| fig-asp: 0.8
#| fig-align: center

@@

```{r}
#| label: fig-skeletal
#| fig-cap: "A) general view of the sample (x250 magnification) showing a fracture and a skeletal remain. B) detail of skeletal remains (x1.3k magnification) composed of calcium (B1), phosphorus (B2) and a small peak of sulphur (B3). The skeletal remains seem to be oxidized, with small peaks of iron within (B4). Geological sample SP16_BLX from the LJW formation."

knitr::include_graphics(here::here('analysis/figures/M3_3_map1.png'))

@@

```{r}
#| label: fig-oval-fossils
#| fig-cap: "A-B) general view of the sample (x190 and x250 magnification, respectively) showing multiple oval fossils and other inclusions. C) detail of an irregular inclusion (x1k magnification) composed of titanium (C1) and an oval skeletal remains composed mainly of calcium (C2), phosphorus (C3) and a small peak of sulphur. Archaeological sample from T4 (H20-4359)."

knitr::include_graphics(here::here('analysis/figures/M15_4.png'))

@@

```{r}
#| label: fig-t6-general
#| fig-cap: "A) general view of the sample (x250 magnification). B) detail of an inclusion seen in A (x3.2k magnification), composed mostly of iron and small peaks of calcium and phosphorus, making this inclusion a possible oxidised skeletal remains of a fossil (1). Archaeological sample from T6 (H20-3890). C) general view of the sample (x250 magnification) with no visible inclusions. D) detail of the sample (x4.2k magnification) showing a small and irregular iron inclusion (2). Archaeological sample from T6 (I19-2846). Dot represents the location of the measured spectra plotted in graphs 1-2."

knitr::include_graphics(here::here('analysis/figures/M18_7_M18_8_graphs.png'))

@@

```{r}
#| label: fig-framboids
#| fig-cap: "A) detail of the sample (x6.5k magnification) showing a possibly spherical pyrite framboid composed of iron (A1) and sulphur (A2). B) detailed view of an inclusion (x1k magnification) composed of iron (B1) and sulphur (B2). Geological sample SP59_JOR_002 from the UJE formation."

knitr::include_graphics(here::here('analysis/figures/M10_5_map1_map2.png'))

@@

```{r}
#| label: fig-calcite
#| fig-cap: "General view of calcite inclusions (light grey) composed mainly of calcium and silica (1). Dot represents the location of the measured spectrum plotted in graph 1. Geological sample SP69_MAR from the UJW formation."

knitr::include_graphics(here::here('analysis/figures/M4_3_graphs.png'))

@@

When comparing the regional Algarve cherts (fig. @fig-pca a), we see that the two first dimensions explain a combined `r 34.6+26.3`% variability in the results, with dimension 1 explaining 34.6% and dimension 2 explaining 26.3%. By observing the scatterplots, it is noticeable that the Lower Jurassic cherts (LJW) are highly variable when compared to the other cherts, with samples which display higher positive values in dimension 1 (Dim1) and dimension 2 (Dim2). However, when focusing on the concentration of LJW results, they seem to fit within the range shown by Middle Jurassic and Upper Jurassic eastern cherts. In comparison, Upper Jurassic cherts from the western basin show higher positive values in dimension 2, which is explained by the higher values of CaO.

The differences between the local Lower Jurassic cherts and non-local cherts are more marked. @fig-pca b represents the correlation between elements and Lower Jurassic cherts, central Portugal Cretaceous cherts and southern Spain Upper Jurassic cherts. Dimension 1 explains \~57% of the variability, while dimension 2 explains \~29%, with a combined percentage of `r 56.6+28.6`%. Once again, it is clear that Jurassic cherts, both from southwestern Portugal and southern Spain, show variability, although they are mostly located in the center/bottom of dimension 2 and center of dimension 1. When compared to these values, central Portugal Cretaceous cherts are located in the negative axis of dimension 1 and show little variability.

A similar pattern emerges when applying a PCA using the same elements but comparing them to archaeological, non-local chert T6, previously attributed to central Portugal (Fig. @fig-pca c). In this PCA, dimension 1 explains \~52% of the variability and dimension 2 \~30%, with a combined percentage of `r 52.2+30.1`%. Similarly to geological Cretaceous samples from central Portugal, T6 samples are also concentrated in the negative side of dimension 1, with a clear overlap between the two.

As observed in @fig-pca b, @fig-pca d shows a concentration of all groups in the centre/bottom of dimension 2 (28.7%) and centre of dimension 1 (58.7%), with the first two dimensions explaining a combined percentage of `r 58.7+28.7`% of the variability. The archaeological local cherts T2 to T5 and non-local type T7 (previously attributed to the Upper Jurassic cherts of southern Spain) are congruent with the corresponding geological sources, although with clear decreased variability.
